
# Cheat sheets

## Scoped verbs vs. purrr

It can be easy to get confused between [purrr](https://dcl-prog.stanford.edu/purrr-basics.html) and scoped verbs. The following diagram illustrates which to use for different combinations of inputs and outputs. For example, use a scoped verb if you want to start and end with a tibble, but purrr if you want to start with a tibble and end up with a vector.

```{r echo=FALSE}
knitr::include_graphics(
  "images/scoped-verbs/purrr-vs-scoped-verbs.png", 
  dpi = image_dpi
)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# Libraries
library(tidyverse)
library(googlesheets4)
library(knitr)
# Parameters
  # Spreadsheet with content
key <- "1TDkJNigZ4c2nVy18N116HwCnkQblRFP3HIkOsmWEo9U"
```

## Suffixes

```{r}
sheets_read(key, sheet = "suffix") %>% 
  select(suffix, `use when`) %>% 
  kable() 
```

## Examples

```{r}
examples <-
  sheets_read(key, sheet = "examples") %>% 
  mutate(
    example_type = if_else(is.na(example_type), lag(example_type), example_type)
  )
```

### `mutate()`, `summarize()`, `select()`, and `rename()`

#### Named functions

```{r}
examples %>% 
  filter(example_type == "named") %>% 
  slice(-n()) %>% 
  select(-example_type) %>% 
  mutate_all(str_replace_na, " ") %>% 
  rename_all(~(str_to_title(.) %>% str_replace("_", " "))) %>% 
  kable(format = "markdown") #have to specify otherwise doesn't handle empty rows well
```

#### Extra arguments

```{r}
examples %>%
  filter(str_detect(example_type, "(extra)")) %>%
  select(-example_type) %>% 
  kable()
```

#### Anonymous functions

```{r}
examples %>% 
  filter(str_detect(example_type, "anonymous")) %>% 
  select(-example_type) %>% 
  kable()
```

### `filter()`

```{r}
examples %>% 
  filter(example_type == "filter") %>% 
  slice(-n()) %>% 
  select(-example_type) %>% 
  mutate_all(str_replace_na, " ") %>% 
  kable(format = "markdown")
```
